Asymptotic statistics of carrier phase and amplitude estimation for PSK using pilots and data

### 2013-11-14 paperandrenew.tex
added andre's version for ref

@@

%opening
\title{Carrier phase and amplitude estimation for phase shift keying using pilots and data}
\author{Robby McKilliam, Andr\'{e} Pollok, Bill Cowley, I. Vaughan L. Clarkson and Barry Quinn  
\thanks{
A preliminary version of this paper has been submitted to ICASSP'13~\cite{McKilliam_leastsqPSKnoncoICASSP_2012}.  Supported under the Australian Government’s Australian Space Research Program.
Robby McKilliam, Andr\'{e} Pollok and Bill Cowley are with the Institute for Telecommunications Research, The University of South Australia, SA, 5095.  Vaughan~Clarkson is with the School of Information Technology \& Electrical Engineering, The University of Queensland, QLD., 4072, Australia.  Barry~Quinn is with the Department of Statistics, Macquarie University, Sydney, NSW, 2109, Australia.
}}

@@

\begin{abstract}
We consider least squares estimators of carrier phase and amplitude from a noisy communications signal that contains both pilot signals, known to the receiver, and data signals, unknown to the receiver.  We focus on signaling constellations that have symbols evenly distributed on the complex unit circle, i.e., $M$-ary phase shift keying.  We show, under reasonably mild conditions on the distribution of the noise, that the least squares estimator of carrier phase is strongly consistent and asymptotically normally distributed.  However, the amplitude estimator is not consistent and converges to a positive real number that is a function of the true carrier amplitude, the noise distribution and the size of the constellation.  %Our theoretical results can also be applied to the case where no pilot symbols exist, i.e., noncoherent detection.  
The results of Monte Carlo simulations are provided and these agree with the theoretical results.   
\end{abstract}
\begin{IEEEkeywords}
Detection, phase shift keying, asymptotic statistics
\end{IEEEkeywords}

@@

In passband communication systems the transmitted signal typically undergoes time offset (delay), phase shift and attenuation (amplitude change).  These effects must be compensated for at the receiver. In this paper we assume that the time offset has been previously handled and we focus on estimating the phase shift and attenuation.  We consider signalling constellations that have symbols evenly distributed on the complex unit circle such as binary phase shift keying (BPSK), quaternary phase shift keying (QPSK) and $M$-ary phase shift keying ($M$-PSK).  In this case, the transmitted symbols take the form,
\[
s_i = e^{j u_i},
\]
where $j = \sqrt{-1}$ and $u_i$ is from the set $\{0, \tfrac{2\pi}{M}, \dots, \tfrac{2\pi(M-1)}{M}\}$ and $M \geq 2$ is the size of the constellation.  We assume that some of the transmitted symbols are \emph{pilot symbols} known to the receiver and the remainder are information carrying \emph{data symbols} with phase that is unknown to the receiver.  So,
\[
s_i = \begin{cases}
p_i & i \in P \\
d_i & i \in D,
\end{cases}
\]
where $P$ is the set of indices describing the position of the pilot symbols $p_i$, and $D$ is a set of indices describing the position of the data symbols $d_i$.  The sets $P$ and $D$ are disjoint, i.e., $P \cap D = \emptyset$  where $\emptyset$ is the empty set, and we let $L = \abs{P \cup D}$ be the total number of symbols transmitted.

We assume that %time offset estimation has been performed and that 
$L$ noisy $M$-PSK symbols are observed by the receiver,  %The received signal after matched filtering is,
\begin{equation}\label{eq:sigmod}
y_i = a_0 s_i + w_i, \qquad i \in P \cup D,
\end{equation}
where $w_i$ is noise and $a_0 = \rho_0 e^{j\theta_0}$ is a complex number representing both carrier phase $\theta_0$ and amplitude $\rho_0$ (by definition $\rho_0$ is a positive real number).  Our aim is to estimate $a_0$ from the noisy symbols $\{ y_i, i \in P \cup D \}$.  Complicating matters is that the data symbols $\{d_i, i \in D\}$ are not known to the receiver and must also be estimated.  Estimation problems of this type have undergone extensive prior study~\cite{ViterbiViterbi_phase_est_1983,Cowley_ref_sym_carr_1998,Wilson1989,Makrakis1990,Liu1991,Mackenthun1994,Sweldens2001,McKilliamLinearTimeBlockPSK2009,Divsalar1990}.  A practical approach is the least squares estimator, that is, the minimisers of the sum of squares function
\begin{equation}\label{eq:SSdefn}
\begin{split}
SS(a, &\{d_i, i \in D\}) = \sum_{i \in P \cup D} \abs{ y_i - a s_i }^2  \\
&= \sum_{i \in P} \abs{ y_i - a p_i }^2 + \sum_{i \in D} \abs{ y_i - a d_i }^2,
\end{split}
\end{equation}
where $\abs{x}$ denotes the magnitude of the complex number $x$.  The least squares estimator is also the maximum likelihood estimator under the assumption that the noise sequence $\{w_i, i \in \ints\}$ is white and Gaussian.  However, as we show, the estimator works well under less stringent assumptions.  %It is the least squares estimator that we primarily study in this paper.

% The existing literature~\cite{Mackenthun1994,Cowley_ref_sym_carr_1998,ViterbiViterbi_phase_est_1983,Sweldens2001,Wilson1989,Makrakis1990,Liu1991} mostly considers what is called \emph{noncoherent detection} (also know as \emph{blind}, \emph{non-data aided} or \emph{non-decision directed detection}) where no pilot symbols exist ($P = \emptyset$).  In the noncoherent setting \emph{differential encoding} is often used, and for this reason the estimation problem has been called \emph{multiple symbol differential detection}.  A popular approach is estimator of Viterbi and Viterbi~\cite{ViterbiViterbi_phase_est_1983}.  The idea is to `strip' the modulation from the received signal by taking $y_i / \abs{y_i}$ to the power of $M$.  A function $F: \reals \mapsto \reals$ is chosen and the estimator of the carrier phase $\theta_0$ is taken to be $\tfrac{1}{M}\angle{C}$ where $\angle$ denotes the complex argument and
% \begin{equation}\label{eq:viterbiviterbi}
% C = \frac{1}{L}\sum_{i \in P \cup D} F(\abs{y_i}) \big(\tfrac{y_i}{\abs{y_i}}\big)^M.
% \end{equation}
% Various choices for $F$ are suggested in~\cite{ViterbiViterbi_phase_est_1983} and a statistical analysis is presented.  A caveat of this estimator is that it is not obvious how pilot symbols should be included. %Techniques are available but they are somewhat ad hoc~\cite{Cowley_ref_sym_carr_1998}.  
% This problem does not occur with the least square estimator.

An important paper is by Mackenthun~\cite{Mackenthun1994} who described an algorithm to compute the least squares estimator requiring only $O(L \log L)$ arithmetic operations.  Sweldens~\cite{Sweldens2001} rediscovered Mackenthun's algorithm in 2001.  Both Mackenthun and Sweldens considered what is called \emph{noncoherent detection} (also know as \emph{blind}, \emph{non-data aided} or \emph{non-decision directed detection}) where no pilot symbols exist ($P = \emptyset$).  We show in Section~\ref{sec:least-squar-estim}~that Mackenthun's algorithm can be modified to include pilot symbols. Our model includes the noncoherent case by setting the number of pilot symbols to zero, that is, putting $P = \emptyset$.

Aside from this extention of Mackenthun's algorithm, the primary purpose of this paper is to analyse the least square estimator of carrier amplitude and phase obtained by minimising~\eqref{eq:SSdefn}.  In the literature, analytic studies of non-data aided estimators of carrier phase have primarily involved the derivation of Cramer-Rao lower bounds under the assumption that the noise $\{w_i\}$ is complex Gaussian distributed~\cite{Delmas_exact_crb_psk_2008,Steendam_lowsnr_crb_2001,Cowley_crbs_phase_freq_1996,DAndrea_modified_bounds_1994}.  The bounds themselves are often complicated to compute and analyse and for this reason it is common to make approximations, known as \emph{modified} Cramer-Rao bounds~\cite{DAndrea_modified_bounds_1994}, or to analyse bound that are valid asymptoticially as the noise variance increases to infinity or decreases to zero~\cite{Moeneclaey_limit_crbs_1998,Steendam_lowsnr_crb_2001,Delmas_exact_crb_psk_2008}.  To the authors knowledge, all of the existing bounds are applicable only in the fully non-data aided case or in the completely data aided case, that is, when either no pilot symbols exist, or when no data symbols exist.  Deriving bounds for the common practical scenario where there is a mixture of pilot and data symbols might be possible, but would further complicate already complicated bounds.

For the case of carrier amplitude the situations is even more complicated.  A simulation study conducted by Pauluzzi~\cite{Pauluzzi2000} suggests that all known non-data aided estimators for carrier amplitude exhibit bias.  The bias is often small when the noise variance is small, but typically becomes significant as the noise variance increases.  Neverthess, Cramer-Rao bound have been derived for \emph{unbiased} estimators of carrier amplitude~\cite{Alagha_crb_snr_bpsk_qpsk_2001,Delmas_exact_crb_psk_2008}.  Unfortunately, these bounds may not be meaningful since known non-data aided amplitude estimators appear to be biased~\cite{Pauluzzi2000}.

In this paper we present an alternative approach to the study of estimators for carrier phase and amplitude.  Rather than compute bounds, we directly study the least squares estimators of carrier phase and amplitude asymptotically as the number of transmitted symbols $L\to \infty$.  We show, under reasonably mild conditions on the distribution of the noise, that the least squares estimator of carrier phase is strongly consistent (asymptotically unbiased) and asymptotically normally distributed.  However, the least squares amplitude estimator is asymptotically biased and converges to a positive real number that is a function of the true carrier amplitude, the noise distribution and the size of the constellation $M$.  To our knowledge, this is the first time an analytic description of the bias or variance of non-data aided amplitude estimator has been obtained.  Our expressions for the asymptotic variance and bias are simple to compute.  %Our results are indirectly supported by the simulation study of Pauluzzi~\cite{Pauluzzi2000}.  We present our our set of simulations supporting our analytic results in Section~\ref{}.

%Two criticisms might be directed at our asymptotic approach when compared to the Cramer-Rao bounds already developed in the literature.  Perhaps the most obvious is that the results apply asymptotically as $L\to\infty$.  The simulation results we present in Section~\ref{} confirm that our asymptotic results accurately model the behaviour of the least squares estimator for modest numbers of symbols, for example $L=256$

One criticism that might be directed at our asymptotic approach when compared to the Cramer-Rao bounds already developed in the literature is that our specific results apply only to the least square estimator.  One response to this criticism is that, because the least squares estimator can be rapidly computed in $O(L\log L)$ operations and because it is the maximum likelihood estimator under the common assumption that the noise is white and Gaussian, then it is likely to be a popular practical choice.  Perhaps an even more convinving response is that, while the specific Theorems~\ref{thm:consistency} and~\ref{thm:normality} that we given apply only to the least square estimtor, the techniques that we develop to derive these results can likely be deployed for the analysis of other estimators.  For example, it would be reasonably straigthforward to extend our results to allow data and pilot symbols with varying constellation size, i.e. varying $M$.  For example, one might give more importance to certain data symbols and use BPSK ($M=2$) for these, but QPSK ($M=4$) for other less important symbols.  This is related to what is called \emph{unequal error protection} in the literature~\cite{Aydinlik_turbo_uep_2008,Sandberg_uep_ldpc_2010} and is commonly used in practical systems.  Perhaps a more involved extension would be the inclusion of carrier frequency as a parameter to be estimated.  The techniques we present, combined with those from~\cite{McKilliamFrequencyEstimationByPhaseUnwrapping2009}, would potentially allow this.

% In the literature it has been common to assume that the data symbols $\{d_i, i \in D\}$ are of primary interest and that the complex amplitude $a_0$ is a nuisance parameter.  The metric of performance is correspondingly the \emph{symbol error rate}, or \emph{bit error rate}.  While estimating the data symbols is ultimately the goal, we take the opposite point of view here.  Our aim is to estimate $a_0$, and we treat the unknown data symbols as nuisance parameters.  This is motivated by the fact that in many modern communication systems the data symbols are \emph{coded}.  For this reason raw symbol error rate is not of interest at this stage.  Instead, we desire an accurate estimator $\hat{a}$ of $a_0$, so that the compensated received symbols $\hat{a}^{-1}y_i$ can be accurately modelled using an additive noise channel.  The additive noise channel is a common assumption for subsequent receiver operations, such as decoding.  The estimator $\hat{a}$ is also used in the computation of decoder metrics for modern decoders, and for interference cancellation in multiuser systems.  Consequently, our metric of performance will not be symbol or bit error rate, but $\sabs{\hat{a} - a_0}$. It will be informative to consider the carrier phase and amplitude estimators separately, that is, if $\hat{a} = \hat{\rho}e^{j\hat{\theta}}$ where $\hat{\rho}$ is a positive real number, then we consider $\sabs{\langle\theta_0 - \hat{\theta}\rangle_\pi}$ and $\sabs{\hat{\rho} - \rho_0}$.  The function $\fracpart{\cdot}_\pi$ denotes its argument taken `modulo $2\pi$' into the interval $[-\pi, \pi)$.  It will become apparent why $\langle\theta_0 - \hat{\theta}\rangle_\pi$ rather than $\theta_0 - \hat{\theta}$ is the appropriate measure of error for the phase parameter.  It is usual to write \emph{estimator} minus \emph{error}, i.e., $\hat{\theta} - \theta_0$ rather than $\theta_0 - \hat{\theta}$.  However, it is notationally convenient to use $\langle\theta_0 - \hat{\theta}\rangle_\pi$ rather than $\langle\hat{\theta} - \theta_0\rangle_\pi$ here.

%It is possible to generalise the results we present here to allow data symbols with varying constellation size, i.e. varying $M$.  For example, one might give more importance to certain data symbols and use BPSK ($M=2$) for these, but QPSK ($M=4$) for other less important symbols.  This is related to what is called \emph{unequal error protection} in the literature~\cite{Aydinlik_turbo_uep_2008,Sandberg_uep_ldpc_2010}.  To keep our ideas and notation focused we don't consider this further here.

The paper is organised in the following way.  Section~\ref{sec:least-squar-estim} extends Mackenthun's algorithm for the coherent case, when both pilot symbols and data symbols exist.  Section~\ref{sec:circ-symm-compl} describes properties of complex random variables that we need.  Section~\ref{sec:stat-prop-least} states two theorems that describe the statistical properties of the least squares estimator of carrier phase $\hat{\theta}$ and amplitude $\hat{\rho}$.  We show, under reasonably general assumptions about the distribution of the noise $w_1,\dots,w_L$, that $\langle\theta_0 - \hat{\theta}\rangle_\pi$ converges almost surely to zero and that $\sqrt{L}\langle\theta_0 - \hat{\theta}\rangle_\pi$ is asymptotically normally distributed as $L\rightarrow \infty$.  However, $\hat{\rho}$ is not a consistent estimator of the amplitude $\rho_0$.  The asymptotic bias of $\hat{\rho}$ is small when the signal to noise ratio (SNR) is large, but the asymptotic bias is significant when the SNR is small.  Sections~\ref{sec:proof-almost-sure} and~\ref{sec:proof-asympt-norm} provide proofs of the theorems stated in Section~\ref{sec:stat-prop-least}.  In Section~\ref{sec:gaussian-noise-case} we consider the special case when the noise is Gaussian.  In this case, our expressions for the asymptotic distribution can be simplified.  Section~\ref{sec:simulations} presents the results of Monte-Carlo simulations.  These simulations agree with the derived asymptotic properties.

@@

\section{Mackenthun's algorithm with pilots}\label{sec:least-squar-estim}

In this section we derive Mackentun's algorithm to compute the least squares estimator of the carrier phase and amplitude~\cite{Mackenthun1994}.  Mackenthun specifically considered the noncoherent setting, so we modify the algorithm to include the pilot symbols.  For the purpose of analysing computational complexity, we will assume that the number of data symbols $\abs{D}$ is proportional to the total number of symbols $L$, so that, for example, $O(L) = O(\abs{D})$.  In this case Mackentun's algorithm requires $O(L \log L)$ arithmetic operations.  This complexity arises from the need to sort a list of $\abs{D}$ elements.  
%We use order notation in the standard way, that is, for functions $h$ and $g$, we write $h(N) = O(g(N))$ to mean that there exists a constant $K > 0$ and a finite $N_0$ such that $h(N) \leq K g(N)$ for all $N > N_0$.

%In Section~\ref{sec:line-time-algor} we will show that a full sort is not neccesary, and that the least squares estimator can be implemented in $O(L)$ operations.

Define the sum of squares function
\begin{align}
SS(a, &\{d_i, i \in D\}) = \sum_{i \in P \cup D} \abs{ y_i - a s_i }^2 \nonumber \\
&= \sum_{i \in P \cup D} \big( \abs{y_i}^2 - a s_i y_i^* - a^* s_i^* y_i + aa^* \big), \label{eq:SS}
\end{align}
where $*$ denotes complex conjugate.  The minimiser of $SS$ with respect to $a$ as a function of $\{d_i, i \in D\}$ is
\begin{equation}\label{eq:hata}
\hat{a}(\{d_i, i \in D\}) = \frac{1}{L} \sum_{i \in P \cup D} y_i s_i^* = \frac{1}{L} Y
\end{equation}
where $L = \abs{P \cup D}$ and to simplify our notation we have put 
\[
Y = \sum_{i \in P \cup D} y_i s_i^* = \sum_{i \in P } y_i p_i^* + \sum_{i \in D } y_i d_i^*.
\]  
Note that $Y$ is a function of the unknown data symbols $\{ d_i, i \in D\}$ and we could write $Y(\{ d_i, i \in D\})$, but have chosen to suppress the argument $(\{ d_i, i \in D\})$ for notational brevity.  Substituting $\frac{1}{L}Y$ for $a$ into~\eqref{eq:SS} we obtain $SS$ minimised with respect to $a$,
\begin{equation}\label{eq:SSdatasymbols}
SS(\{d_i, i \in D\}) = A - \frac{1}{L}\abs{Y}^2,
\end{equation}
where $A = \sum_{i \in P \cup D}\abs{y_i}^2$ does not depend on the $d_i$.  The least squares estimators of the data symbols are the minimisers of~\eqref{eq:SSdatasymbols}.  Observe that given candidate values for the data symbols, we can compute the corresponding $SS(\{d_i, i \in D\})$ in $O(L)$ arithmetic operations.  It turns out that there are at most $M\abs{D}$ candidate values of the least squares estimator of the data symbols~\cite{Sweldens2001,Mackenthun1994}.  %When the number of data symbols is not small, this set is substantially smaller than the entire set of possible transmitted symbols, which is of size $M^{\abs{D}}$.

>>>>

\usepackage{color}
\newcommand{\comment}[1]{\textcolor{red}{#1}}

@@

% use italics version of capital Greek letters
\renewcommand{\Phi}{\varPhi}
\renewcommand{\Theta}{\varTheta}

%opening
\title{Asymptotic statistics of carrier phase and amplitude estimation for PSK using pilots and data}
\author{Robby McKilliam, Andr\'{e} Pollok, Bill Cowley, I. Vaughan L. Clarkson and Barry Quinn  
\thanks{
A preliminary version of this paper was presented at ICASSP'13~\cite{McKilliam_leastsqPSKnoncoICASSP_2012}.  Supported under the Australian Government's Australian Space Research Program.
Robby McKilliam, Andr\'{e} Pollok and Bill Cowley are with the Institute for Telecommunications Research, The University of South Australia, SA, 5095.  Vaughan~Clarkson is with the School of Information Technology \& Electrical Engineering, The University of Queensland, QLD., 4072, Australia.  Barry~Quinn is with the Department of Statistics, Macquarie University, Sydney, NSW, 2109, Australia.
}}

@@

\begin{abstract}
We consider least squares estimators of carrier phase and amplitude from a noisy communications signal that contains both pilot signals, known to the receiver, and data signals, unknown to the receiver.  We focus on signaling constellations that have symbols evenly distributed on the complex unit circle, i.e., $M$-ary phase shift keying.  We show, under reasonably mild conditions on the distribution of the noise, that the least squares estimator of carrier phase is strongly consistent and asymptotically normally distributed.  However, the amplitude estimator is not consistent and converges to a positive real number that is a function of the true carrier amplitude, the noise distribution and the size of the constellation.  %Our theoretical results can also be applied to the case where no pilot symbols exist.  
The results of Monte-Carlo simulations are provided and these agree with the theoretical results.   
\end{abstract}
\begin{IEEEkeywords}
Detection, phase shift keying, asymptotic statistics
\end{IEEEkeywords}

@@

In passband communication systems the transmitted signal typically undergoes time offset (delay), phase shift and attenuation (amplitude change).  These effects must be compensated for at the receiver. In this paper we assume that the time offset has been previously handled, and we focus on estimating the phase shift and attenuation.  We consider signalling constellations that have symbols evenly distributed on the complex unit circle such as binary phase shift keying (BPSK), quaternary phase shift keying (QPSK) and $M$-ary phase shift keying ($M$-PSK).  In this case, the transmitted symbols take the form,
\[
s_i = e^{j u_i},
\]
where $j = \sqrt{-1}$ and $u_i$ is from the set $\{0, \tfrac{2\pi}{M}, \dots, \tfrac{2\pi(M-1)}{M}\}$ and $M \geq 2$ is the size of the constellation.  We assume that some of the transmitted symbols are \emph{pilot symbols} known to the receiver and the remainder are information carrying \emph{data symbols} with phase that is unknown to the receiver.  So,
\[
s_i = \begin{cases}
p_i, & i \in P \\
d_i, & i \in D,
\end{cases}
\]
where $P$ is the set of indices describing the position of the pilot symbols $p_i$, and $D$ is a set of indices describing the position of the data symbols $d_i$.  The sets $P$ and $D$ are disjoint, i.e., $P \cap D = \emptyset$, where $\emptyset$ is the empty set, and we let $L = \abs{P \cup D}$ be the total number of symbols transmitted.

We assume that %time offset estimation has been performed and that 
$L$ noisy $M$-PSK symbols are observed by the receiver,  %The received signal after matched filtering is,
\begin{equation}\label{eq:sigmod}
y_i = a_0 s_i + w_i, \qquad i \in P \cup D,
\end{equation}
where $w_i$ is noise and $a_0 = \rho_0 e^{j\theta_0}$ is a complex number representing both carrier phase $\theta_0$ and amplitude $\rho_0$ (by definition $\rho_0$ is a positive real number).  Our aim is to estimate $a_0$ from the noisy symbols $\{ y_i, i \in P \cup D \}$.  Complicating matters is that the data symbols $\{d_i, i \in D\}$ are not known to the receiver and must also be estimated.  Estimation problems of this type have undergone extensive prior study~\cite{ViterbiViterbi_phase_est_1983,Cowley_ref_sym_carr_1998,Wilson1989,Makrakis1990,Liu1991,Mackenthun1994,Sweldens2001,McKilliamLinearTimeBlockPSK2009,Divsalar1990}.  A practical approach is the least squares estimator, that is, the minimisers of the sum of squares function
\begin{equation}\label{eq:SSdefn}
\begin{split}
SS(a, &\{d_i, i \in D\}) = \sum_{i \in P \cup D} \abs{ y_i - a s_i }^2  \\
&= \sum_{i \in P} \abs{ y_i - a p_i }^2 + \sum_{i \in D} \abs{ y_i - a d_i }^2,
\end{split}
\end{equation}
where $\abs{x}$ denotes the magnitude of the complex number $x$.  The least squares estimator is also the maximum likelihood estimator under the assumption that the noise sequence $\{w_i, i \in \ints\}$ is additive white and Gaussian.  However, as we show, the estimator works well under less stringent assumptions.  %It is the least squares estimator that we primarily study in this paper.

\comment{
Doesn't Bill's paper \cite{Cowley_ref_sym_carr_1998} consider the case with pilots?
}

\comment{
The existing literature~\cite{Mackenthun1994,Cowley_ref_sym_carr_1998,ViterbiViterbi_phase_est_1983,Sweldens2001,Wilson1989,Makrakis1990,Liu1991} mostly considers \emph{non-data aided} estimation, where no pilot symbols exist ($P = \emptyset$). In this setting \emph{differential encoding} and \emph{differential (i.e. noncoherent) detection} are often used to handle the resulting $M$-fold ambiguity of the phase estimate.  A popular non-data aided estimator was proposed by Viterbi and Viterbi~\cite{ViterbiViterbi_phase_est_1983}.%
}
The idea is to `strip' the modulation from the received signal by taking $y_i / \abs{y_i}$ to the power of $M$.  A function $F: \reals \mapsto \reals$ is chosen and the estimator of the carrier phase $\theta_0$ is taken to be $\tfrac{1}{M}\angle{C}$ where $\angle$ denotes the complex argument and
\begin{equation}\label{eq:viterbiviterbi}
C = \frac{1}{L}\sum_{i \in P \cup D} F(\abs{y_i}) \big(\tfrac{y_i}{\abs{y_i}}\big)^M.
\end{equation}
Various choices for $F$ are suggested in~\cite{ViterbiViterbi_phase_est_1983} and a statistical analysis is presented.  A caveat of this estimator is that it is not obvious how pilot symbols should be included. %Techniques are available but they are somewhat ad hoc~\cite{Cowley_ref_sym_carr_1998}.  
This problem does not occur with the least squares estimator.

An important paper is by Mackenthun~\cite{Mackenthun1994} who described an algorithm to compute the least squares estimator requiring only $O(L \log L)$ arithmetic operations.  Sweldens~\cite{Sweldens2001} rediscovered Mackenthun's algorithm in 2001.  Both Mackenthun and Sweldens considered only the non-data aided setting, but we show in Section~\ref{sec:least-squar-estim}~that Mackenthun's algorithm can be modified to include pilot symbols. Our model includes the non-data aided case by setting the number of pilot symbols to zero, that is, putting $P = \emptyset$.

In the literature it has been common to assume that the data symbols $\{d_i, i \in D\}$ are of primary interest and that the complex amplitude $a_0$ is a nuisance parameter.  The metric of performance is correspondingly the \emph{symbol error rate}, or \emph{bit error rate}.  While estimating the data symbols is ultimately the goal, we take the opposite point of view here.  Our aim is to estimate $a_0$, and we treat the unknown data symbols as nuisance parameters.  This is motivated by the fact that in many modern communication systems the data symbols are \emph{coded}.  For this reason raw symbol error rate is not of interest at this stage.  Instead, we desire an accurate estimator $\hat{a}$ of $a_0$, so that the compensated received symbols $\hat{a}^{-1}y_i$ can be accurately modelled using an additive noise channel.  The additive noise channel is a common assumption for subsequent receiver operations, such as decoding.  The estimator $\hat{a}$ is also used in the computation of decoder metrics for modern decoders, and for interference cancellation in multiuser systems.  \comment{Furthermore, iterative receiver synchronisation methods that are aided by the error correcting code, also known as \emph{turbo synchronisation} \cite{Lottici2004,Noels2005}, %\cite{Herzet2007,Herzet2007_TurboSyncFramework}
typically converge correctly only if initialised at some $a$ sufficiently close to $a_0$. Obtaining an accurate estimator $\hat{a}$ of $a_0$ is therefore a crucial ingredient for an effective turbo synchronisation receiver.} Consequently, our primary metric of performance will not be symbol or bit error rate, but the estimation error.  \comment{However, we will also present packet error rate results for an LDPC-coded satellite communication system to demonstrate the use of the proposed estimator in a practical system\dots}
%It will be informative to consider the carrier phase and amplitude estimators separately, that is, if $\hat{a} = \hat{\rho}e^{j\hat{\theta}}$ where $\hat{\rho}$ is a positive real number, then we consider $\sabs{\langle\theta_0 - \hat{\theta}\rangle_\pi}$ and $\sabs{\hat{\rho} - \rho_0}$.  The function $\fracpart{\cdot}_\pi$ denotes its argument taken `modulo $2\pi$' into the interval $[-\pi, \pi)$.  It will become apparent why $\langle\theta_0 - \hat{\theta}\rangle_\pi$ rather than $\theta_0 - \hat{\theta}$ is the appropriate measure of error for the phase parameter.  It is usual to write \emph{estimator} minus \emph{truth}, i.e., $\hat{\theta} - \theta_0$ rather than $\theta_0 - \hat{\theta}$.  However, it is notationally convenient to use $\langle\theta_0 - \hat{\theta}\rangle_\pi$ rather than $\langle\hat{\theta} - \theta_0\rangle_\pi$ here.

%It is possible to generalise the results we present here to allow data symbols with varying constellation size, i.e.\ varying $M$.  For example, one might give more importance to certain data symbols and use BPSK ($M=2$) for these, but QPSK ($M=4$) for other less important symbols.  This is related to what is called \emph{unequal error protection} in the literature~\cite{Aydinlik_turbo_uep_2008,Sandberg_uep_ldpc_2010}.  To keep our ideas and notation focused we don't consider this further here.

The paper is organised in the following way.  Section~\ref{sec:least-squar-estim} extends Mackenthun's algorithm for the case, when both pilot symbols and data symbols exist.  Section~\ref{sec:stat-prop-least} states two theorems that describe the asymptotic statistical properties of the least squares estimator of carrier phase and amplitude.
%Section~\ref{sec:stat-prop-least} states two theorems that describe the statistical properties of the least squares estimator of carrier phase $\hat{\theta}$ and amplitude $\hat{\rho}$.  We show, under reasonably general assumptions about the distribution of the noise $w_1,\dots,w_L$, that $\langle\theta_0 - \hat{\theta}\rangle_\pi$ converges almost surely to zero and that $\sqrt{L}\langle\theta_0 - \hat{\theta}\rangle_\pi$ is asymptotically normally distributed as $L\rightarrow \infty$.  However, $\hat{\rho}$ is not a consistent estimator of the amplitude $\rho_0$.  The asymptotic bias of $\hat{\rho}$ is small when the signal to noise ratio (SNR) is large, but significant when the SNR is small.  
Sections~\ref{sec:proof-almost-sure} and~\ref{sec:proof-asympt-norm} provide proofs of the theorems stated in Section~\ref{sec:stat-prop-least}.  
%In Section~\ref{sec:gaussian-noise-case} we consider the special case when the noise is Gaussian.  In this case, our expressions for the asymptotic distribution can be simplified.  
In Section~\ref{sec:simulations} we present the results of Monte-Carlo simulations.  These simulations agree with the derived asymptotic properties.

@@

\comment{
\section{Mackenthun's algorithm with pilots}\label{sec:least-squar-estim}
}

%In this section we derive Mackentun's algorithm to compute the least squares estimator of the carrier phase and amplitude~\cite{Mackenthun1994}.  
Mackenthun specifically considered the non-data aided setting and assumed differential encoding and detection to deal with the $M$-fold phase ambiguity \cite{Mackenthun1994}. In contrast, we consider non-differential encoding with coherent detection and to this end, extend Mackenthun's algorithm to include pilot symbols.

For the purpose of analysing computational complexity, we will assume that the number of data symbols $\abs{D}$ is proportional to the total number of symbols $L$, so that, for example, $O(L) = O(\abs{D})$.  In this case Mackentun's algorithm requires $O(L \log L)$ arithmetic operations.  This complexity arises from the need to sort a list of $\abs{D}$ elements.  
%We use order notation in the standard way, that is, for functions $h$ and $g$, we write $h(N) = O(g(N))$ to mean that there exists a constant $K > 0$ and a finite $N_0$ such that $h(N) \leq K g(N)$ for all $N > N_0$.

%In Section~\ref{sec:line-time-algor} we will show that a full sort is not necessary, and that the least squares estimator can be implemented in $O(L)$ operations.

Our starting point is the sum of squares function defined in \eqref{eq:SSdefn}, which can be rewritten as
%Define the sum of squares function
\begin{align}
SS(a, &\{d_i, i \in D\}) = \sum_{i \in P \cup D} \abs{ y_i - a s_i }^2 \nonumber \\
&= \sum_{i \in P \cup D} \big( \abs{y_i}^2 - a s_i y_i^* - a^* s_i^* y_i + aa^* \big), \label{eq:SS}
\end{align}
where $*$ denotes complex conjugation.  The minimiser of $SS$ with respect to $a$ as a function of $\{d_i, i \in D\}$ is
\begin{equation}\label{eq:hata}
\hat{a}(\{d_i, i \in D\}) = \frac{1}{L} \sum_{i \in P \cup D} y_i s_i^* = \frac{1}{L} Y(\{d_i, i \in D\}),
\end{equation}
where $L = \abs{P \cup D}$ and to simplify our notation we have put 
\begin{align}\label{eq:Y}
	Y(\{d_i, i \in D\}) = \sum_{i \in P \cup D} y_i s_i^* 
		%= \sum_{i \in P } y_i p_i^* + \sum_{i \in D } y_i d_i^*
		= B + \sum_{i \in D } y_i d_i^*,
\end{align}
where $B = \sum_{i \in P} y_i p_i^*$ is independent of the data symbols.  For notational brevity, we also write $Y$ instead of $Y(\{ d_i, i \in D\})$.  Substituting $\hat{a}$ from \eqref{eq:hata} for $a$ into~\eqref{eq:SS} we obtain $SS$ minimised with respect to $a$,
\begin{equation}\label{eq:SSdatasymbols}
	SS(\{d_i, i \in D\}) = A - \frac{1}{L}\abs{Y(\{d_i, i \in D\})}^2,
\end{equation}
where $A = \sum_{i \in P \cup D}\abs{y_i}^2$ does not depend on the $d_i$.  The least squares estimators of the data symbols, denoted $\{\hat{d}_i,i\in D\}$, are the minimisers of~\eqref{eq:SSdatasymbols} or, equivalently, the maximisers of $\abs{Y}^2$.
%\begin{align}
%	Q(\{d_i, i \in D\}) = \abs{Y(\{d_i, i \in D\})}^2.
%\end{align}
Observe that given candidate values for the data symbols, we can compute the corresponding $\abs{Y}^2$ in $O(L)$ arithmetic operations.  The na\"{\i}ve approach is to compute the objective function $\abs{Y}^2$ for all possible $M^\abs{D}$ data sequences and select the sequence that yields the maximum value.  Using Mackenthun's algorithm, we are able to avoid this computationally expensive exhaustive search, which would require $O(LM^L)$ arithmetic operations.

First consider the case without pilot symbols, i.e.\ $P=\emptyset$. In this case, $\abs{Y}^2$ is identical to the objective function considered by Mackenthun \cite[Eq.\ (2)]{Mackenthun1994}.  This equivalence stems from the fact that the least squares estimators $\hat{d}_i$ are also the maximum likelihood estimators under the additive white Gaussian noise assumption of \cite{Mackenthun1994}.  Consequently, Mackenthun's algorithm as presented in~\cite{Mackenthun1994} or~\cite{Sweldens2001} can be applied without modification to obtain the $\hat{d}_i$.  As shown in~\cite{Sweldens2001,Mackenthun1994}, at most $M\abs{D}$ candidate sequences need to be trialled. In fact, due to an $M$-fold phase ambiguity in the absence of pilot symbols, the objective function $\abs{Y}^2$ is periodic in $\abs{D}$ and thus, only $\abs{D}$ of the $M\abs{D}$ candidate sequences need to be considered. We do not explicitly state the algorithm at this point, but instead describe it below for the more general setting with pilots.

In the presence of pilot symbols, all $M\abs{D}$ candidate sequences are relevant and we rely on the pilots to resolve the phase ambiguity. In the following we describe how these sequences can be enumerated. Let $\hat{b}_i$ denote the hard decision on the received data symbol $y_i$, $i \in D$, i.e.
\[
	\hat{b}_i = e^{j\hat{u}_i}, \qquad \text{where} \qquad \hat{u}_i = \round{\angle y_i},
\]
where $\angle(\cdot)$ denotes the complex argument (or phase), and $\round{\cdot}$ rounds its argument to the nearest multiple of $\frac{2\pi}{M}$.  A word of caution, the notation $\round{\cdot}$ is often used to denote rounding to the nearest \emph{integer}.  This is not the case here.  If the function $\operatorname{round}(\cdot)$ takes its argument to the nearest integer\footnote{The direction of rounding for half-integers is not important so long as it is consistent.  We have chosen to round up half-integers here.} then,
\[
\round{x} = \tfrac{2\pi}{M}\operatorname{round}\left(\tfrac{M}{2\pi}x\right).
\]
